Computer-Assisted Analysis of the \\ Anderson-Hájek Ontological Controversy

### 2017-01-06 Papers/2017/LogicaUniversalis-HajekAndersonControversy/main.tex
Improve various sentences

In preceding experiments \citep{C55,C40,J30}, we have already
demonstrated that the technology we employ is well suited for the
task. In fact, computers may even contribute philosophically relevant
new knowledge. For example, the theorem prover Leo-II \cite{Leo-II}
detected an inconsistency in Gödel's original script
\citep{GoedelNotes} of the argument, which was undetected by
philosophers before.  This inconsistency obviously renders Gödel's
original version of the argument as pointless.  However, as our
preceding experiments also confirmed, Scott's variant
\citep{ScottNotes} from Fig.~\ref{fig:Goedel} is immune to this
issue.\footnote{Gödel avoids the first conjunct $\varphi(x)$ in the
  definition of \emph{essence (D2)}, which Scott added for cosmetic
  reason. However, as Leo-II detected, one can prove from Gödel's
  definition that the empty property becomes an essence of every
  individual, which, together with theorem T1, axiom A5 and the
  defintion of \emph{necessary existence (D3)} causes the
  inconsistency. For more details see \citet{C55}.} Hence, in the
remainder of this article the term \emph{Gödel's ontological argument}
actually refers to Scott's variant of it, whose soundness has been
formally 
verified with our technology (for logics entailing higher-order modal
logic KB with constant domain or varying domain semantics). We here
assume some familiarity with these related experiments and papers.

@@

The particular motivation and context for the work presented in the given
article is as follows: It is well known that the axioms in G\"odel's
ontological proof (cf. Fig. \ref{fig:Goedel}) entail what is called
\emph{modal collapse} \citep{Sobel1987,SobelBook2004}. This means that
the formula $\varphi \rightarrow \Box \varphi$, abbreviated as MC,
holds for any formula $\varphi$ and not just for
$\exists x. \mathit{God}(x)$ as intended.  Hence, anything that is the case is necessarily the case, actual truth implies necessary truth, or, in other words, there are no
contingent truths. In fact, one may even go further and interpret MC as a result
against free will.
%
The fact that MC follows from Gödel's axioms, which has also been
confirmed with higher-order automated theorem provers in our previous
experiments \citep{C40,J30}, has led to strong criticism of the
argument and stimulated attempts to remedy the problem.
\citet{Hajek_Magari_and_others_1996,Hajek_der_Mathematiker_2001}
proposed the use of cautious instead of full comprehension principles,
and \citet{fitting02:_types_tableaus_god} took greater care of the
semantics of higher-order quantifiers in the presence of
modalities. Others, such as
\citet{anderson90:_some_emend_of_goedel_ontol_proof},
\citet{Hajek2002} and \citet{bjordal99}, proposed emendations of
G\"odel's axioms and definitions. They require neither comprehension
restrictions nor more complex semantics and are therefore technically
simpler to analyze with computer support. Hence, we here concentrate
on the latter emendations of Gödel's argument and leave the former
ones for future work.  Again, familiarity with the papers by
\citet{anderson90:_some_emend_of_goedel_ontol_proof},
\citet{Hajek2002} and \citet{bjordal99} is assumed, since we here are only
briefly commenting/assessing various claims made in there.

Our formalizations employ the embedding of higher-order modal logic
(HOML) in classical higher-order logic (HOL) as introduced in previous
work \citep{C40,J23} as enabling technology.  We explored the effect
of different domain conditions on the provability of lemmas, theorems
and even axioms.  This was motivated by a controversy between Hájek
and Anderson regarding the redundancy of some axioms in Anderson's
emendation. In \emph{constant domain semantics} (\emph{possibilist
  notion of quantification}), the individual domains are the same in
all possible worlds. In \emph{varying domain semantics}
(\emph{actualist notion of quantification}), the domains may vary from
world to world. The latter notion is technically encoded in our
approach with the help of an existence relation expressing which
individuals actually exist in each world. In other words, actualist
quantification is formalized as possibilist quantification guarded by
the existence relation. This technical solution
enables us to flexibly switch between these different notions of
quantification and to even mix them. When we mention varying
domain semantics (actualist quantification) below, we in fact mean
varying domain semantics for the domain of individuals only, while for
the other types we still assume the standard constant domain semantics
(possibilist quantification).  This is, to our best knowledge, in line 
with the intention of the authors (it would of course be easily
possible in our approach to study other combinations).

In our
experiments we have utilized the proof assistant Isabelle/HOL
\citep{Isabelle} together with the external automated higher-order provers
Leo-II \citep{Leo-II} and Satallax \citep{Satallax}, called through Sledgehammer. Sledgehammer analyzes the proofs generated by these provers and tells which axioms or previously proven lemmas are need to reconstruct the proof inside Isabelle using Isabelle's \emph{metis} method \citep{PaulsonSusanto}, which relies on a Isabelle porting of the resolution-based first-order prover Metis
\citep{Hurd03first-orderproof}. To find counter-examples, we used Nitpick \citep{Nitpick}, also through Isabelle.

@@

Next, we address some controversially debated claims regarding the redundancy,
superfluousness and independence of particular axioms in the
investigated emendations and variants. Several novel findings are
reported.

>>>>

In preceding experiments \citep{C55,C40,J30}, we have already
demonstrated that the technology we employ is well suited for the
task. In fact, computers may even contribute philosophically relevant
new knowledge. For example, the theorem prover Leo-II \cite{Leo-II}
detected an inconsistency in Gödel's original script
\citep{GoedelNotes} of the argument, which was undetected by
philosophers before.  This inconsistency renders pointless Gödel's
original version of the argument. Nevertheless, as our
preceding experiments also confirmed, Scott's variant
\citep{ScottNotes} from Fig.~\ref{fig:Goedel} is immune to this
issue.\footnote{Gödel avoids the first conjunct $\varphi(x)$ in the
  definition of \emph{essence (D2)}, which Scott added for cosmetic
  reason. However, as Leo-II detected, one can prove from Gödel's
  definition that the empty property becomes an essence of every
  individual, which, together with theorem T1, axiom A5 and the
  defintion of \emph{necessary existence (D3)} causes the
  inconsistency. For more details see \citet{C55}.} Hence, in the
remainder of this article the term \emph{Gödel's ontological argument}
actually refers to Scott's variant, whose logical validity has been
formally 
verified with our technology (for higher-order modal
logic KB (or stronger logics) with constant domain or varying domain semantics). Here we
assume some familiarity with these related experiments and papers.

@@

The particular motivation and context for the work presented in the given
article is as follows: It is well known that the axioms in G\"odel's
ontological proof (cf. Fig. \ref{fig:Goedel}) entail what is called
\emph{modal collapse} \citep{Sobel1987,SobelBook2004}. This means that
the formula $\varphi \rightarrow \Box \varphi$, abbreviated as MC,
holds for any formula $\varphi$ and not just for
$\exists x. \mathit{God}(x)$ as intended.  Hence, anything that is the case is necessarily the case, actual truth implies necessary truth, or, in other words, there are no
contingent truths. In fact, one may even go further and interpret MC as a result
against free will.
%
The fact that MC follows from Gödel's axioms, which has also been
confirmed with higher-order automated theorem provers in our previous
experiments \citep{C40,J30}, has led to strong criticism of the
argument and stimulated attempts to remedy the problem.
\citet{Hajek_Magari_and_others_1996,Hajek_der_Mathematiker_2001}
proposed the use of cautious restrictions of comprehension principles,
and \citet{fitting02:_types_tableaus_god} took greater care of the
semantics of higher-order quantifiers in the presence of
modalities. Others, such as
\citet{anderson90:_some_emend_of_goedel_ontol_proof},
\citet{Hajek2002} and \citet{bjordal99}, proposed emendations of
G\"odel's axioms and definitions. They require neither comprehension
restrictions nor more complex semantics and are therefore technically
simpler to analyze with computer support. Therefore, we concentrate
on the latter emendations of Gödel's argument and leave the former
ones for future work.  Again, familiarity with the papers by
\citet{anderson90:_some_emend_of_goedel_ontol_proof},
\citet{Hajek2002} and \citet{bjordal99} is assumed, since we only
briefly assess various claims made there.

Our formalizations employ the embedding of higher-order modal logic
(HOML) in classical higher-order logic (HOL) as introduced in previous
work \citep{C40,J23} as an enabling technology to settle a long-standing
controversy between Hájek
and Anderson regarding the redundancy of some axioms in Anderson's
emendation. A core issue in the controversy was whether Anderson's emendation
should be interpreted with constant or varying domains and how the provability
of lemmas, theorems and even axioms was affected by this choice.
In \emph{constant domain semantics} (\emph{possibilist
  notion of quantification}), the individual domains are the same in
all possible worlds. In \emph{varying domain semantics}
(\emph{actualist notion of quantification}), the domains may vary from
world to world. The latter notion is technically encoded in our
approach with the help of an existence relation expressing which
individuals actually exist in each world. In other words, actualist
quantification is formalized as possibilist quantification guarded by
the existence relation. This technical solution
enables us to flexibly switch between these different notions of
quantification and to even mix them. When we mention varying
domain semantics (actualist quantification) below, we in fact mean
varying domain semantics for the domain of individuals only, while for
the other types we still assume the standard constant domain semantics
(possibilist quantification).  This is, to our best knowledge, in line 
with the intention of the authors (it would of course be easily
possible in our approach to study other combinations).

In our
experiments we have utilized the proof assistant Isabelle/HOL
\citep{Isabelle} together with the external automated higher-order provers
Leo-II \citep{Leo-II} and Satallax \citep{Satallax}, called through Sledgehammer. Sledgehammer analyzes the proofs generated by these provers and tells which axioms or previously proven lemmas are needed to reconstruct the proof inside Isabelle using its \emph{metis} method \citep{PaulsonSusanto}, which relies on a reimplementation of the resolution-based first-order prover Metis
\citep{Hurd03first-orderproof}. To find counter-examples, we used Nitpick \citep{Nitpick}, also through Isabelle.

@@

Next, we investigate some controversially debated claims regarding the redundancy,
superfluousness and independence of particular axioms in the
investigated emendations and variants. Several novel findings are
reported.
